---
title: "Getting started with nva"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with nva}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

The nva package provides an R interface to the [Norwegian National Research
Archive (NVA)](https://nva.sikt.no) API. It covers both the NVA publication
search and the Cristin registry for persons, organizations, and projects.

## Installation

```{r}
# Install from GitHub
# install.packages("pak")
pak::pak("roarst/nva")
```

## Authentication

Most NVA endpoints are public and do not require authentication. An API key is
only needed for endpoints that access non-public data.
If you have an API key, set it for the current session:

```{r}
library(nva)

nva_set_api_key("your-api-key")

# Check if a key is configured
nva_has_api_key()
```

For persistent configuration, add `NVA_API_KEY=your_key` to your `.Renviron`
file (use `usethis::edit_r_environ()` to open it).

## Searching publications

`nva_search()` is the main entry point for finding publications. It returns a
tibble with key metadata.

```{r}
# Basic search
nva_search("climate change")

# Filter by year and limit results
nva_search("machine learning", year = 2024, limit = 20)

# Filter by organization (Cristin ID, e.g. "185" for University of Oslo)
nva_search("biodiversity", organization = "185")

# Filter by publication type
nva_search("deep learning", type = "AcademicArticle")

# Combine multiple filters
nva_search(
  "renewable energy",
  year = "2020,2024",

  organization = "194",
  scientific_value = "LevelTwo",
  files = "hasPublicFiles"
)
```

The result is a tibble with columns: `identifier`, `title`, `type`, `year`,
`status`, `contributors` (list), and `institutions` (list).

### Search aggregations

Use `nva_search_aggregations()` to see how results are distributed across
categories like publication type, institution, or year — without fetching the
publications themselves.

```{r}
aggs <- nva_search_aggregations("climate change")
aggs

# Filter to see only publication type counts
aggs[aggs$aggregation == "type", ]
```

## Publication details and files

Once you have a publication identifier from search results, retrieve the full
record or its associated files.

```{r}
# Get the full record (returns a list)
pub <- nva_publication("01907b56-a6b0-7b8c-8f79-12345abcdef")
pub$entityDescription$mainTitle

# Get multiple publications as a tibble
pubs <- nva_publications(c("id-1", "id-2"))

# List files attached to a publication
files <- nva_publication_files("01907b56-a6b0-7b8c-8f79-12345abcdef")
files

# Download a file
nva_download_file(files$identifier[1], destfile = "paper.pdf")
```

## Cristin: persons

The Cristin registry contains information about researchers at Norwegian
institutions.

```{r}
# Search for a researcher by name
nva_cristin_person_search(query = "Hansen")

# Search within a specific organization
nva_cristin_person_search(query = "Olsen", organization = "185")

# Get detailed info for a specific person (returns a list)
person <- nva_cristin_person(12345)

# Get structured details for one or more persons as a tibble
nva_cristin_persons(c(12345, 67890))

# Get a researcher's publications
nva_cristin_person_publications(12345)
nva_cristin_person_publications(12345, year = 2024)
```

## Cristin: organizations

Look up Norwegian research institutions and their organizational hierarchy.

```{r}
# Search for organizations
nva_cristin_organization_search("university")

# Get a specific organization (returns a list)
org <- nva_cristin_organization(185)
org$labels$en
# [1] "University of Oslo"

# Get multiple organizations as a tibble
nva_cristin_organizations(c(185, 194))

# Explore the organizational hierarchy
faculties <- nva_cristin_organization_subunits(185)
faculties

# Go deeper (faculties and departments)
nva_cristin_organization_subunits(185, depth = 2)

# Get publications for an organization
nva_cristin_organization_publications(185, year = 2024)
```

## Cristin: projects

Search and retrieve research projects registered in Cristin.

```{r}
# Search for projects
nva_cristin_project_search(query = "climate")

# Filter by status
nva_cristin_project_search(status = "ACTIVE", organization = "185")

# Get a specific project (returns a list)
project <- nva_cristin_project(123456)

# Get multiple projects as a tibble
nva_cristin_projects(c(123456, 789012))
```

## Pagination with fetch_all

Search functions return a limited number of results by default (`limit = 10`).
For larger result sets, use `fetch_all = TRUE` to automatically paginate
through all matching records.

```{r}
# Fetch all matching publications (may be slow for broad queries)
all_pubs <- nva_search("biodiversity", organization = "185", fetch_all = TRUE)

# Cap the number of results
capped <- nva_search(
  "machine learning",
  fetch_all = TRUE,
  max_results = 500
)

# Also works for person and organization publications
all_person_pubs <- nva_cristin_person_publications(12345, fetch_all = TRUE)
```

For Cristin search endpoints (persons, organizations, projects), use the
`page` parameter for manual pagination:

```{r}
page1 <- nva_cristin_person_search(query = "Hansen", limit = 100, page = 1)
page2 <- nva_cristin_person_search(query = "Hansen", limit = 100, page = 2)
```
